High-Dimensional Regularized Discriminant Analysis

### 2015-10-26 reports/paper-hdrda.Rnw
Rewrote the design for the simulated data sets.

In discriminant analysis we wish to assign an unlabeled vector $\bm x \in
\mathbb{R}_{p \times 1}$ to one of $K$ unique, known classes by constructing a
classifier from $N$ training observations. Let $\bm x_i = (x_{i1}, \ldots,
x_{ip}) \in \mathbb{R}_{p \times 1}$ be the $i$th observation $(i = 1, \ldots,
N)$ with true, unique membership $y_i \in \{\omega_1, \ldots, \omega_K\}$. We
assume that $(\bm x_i, y_i)$ is a realization from a mixture distribution $p(\bm
x) = \sum_{k=1}^K p(\bm x | \omega_k) p(\omega_k)$, where $p(\bm x | \omega_k)$
is the probability density function of the $k$th class and $p(\omega_k)$ is the
prior probability of class membership of the $k$th class. We further assume that
each of $p(\omega_k)$ is equal, $k = 1, \ldots, K$.

@@

We compared the \emph{HDRDA} classifer to the five competing classifiers in a
simulaton study. We simulated $K = 3$ $p$-dimensional multivariate normal
populations with a common positive-definite covariance matrix having a
block-diagonal covariance structure, which is intended to resemble
gene-expression data. That is, within each block of pathways, genes are
positively correlated and that this correlation decays as a function of the
distance between any two genes. Thus, we simulated $K = 3$ multivariate normal
populations with the common positive-definite covariance matrix

\begin{align*}
  \bm\Sigma = \bm\Sigma^{(\rho)} \oplus \bm\Sigma^{(-\rho)} \oplus \ldots \oplus \bm\Sigma^{(\rho)},
\end{align*}

where the $(i,j)$th entry of $\bm\Sigma^{(\rho)} \in \mathbb{R}_{25 \times 25}$ is

\begin{align*}
\bm\Sigma_{ij}^{(\rho)} = \{ \rho^{|i - j|} \}_{1 \le i,j \le 25},
\end{align*}

altering positive and negative values of $\rho$ are used from block to block. In
our simulation study, we considered three different values of $\rho$, 0.1, 0.5,
and 0.9, to explore the effect of autocorrelation on classification accuracy,
especially given that the classifiers from \cite{Tong:2012hw} and
\cite{Pang:2009ik} assume independence among features.

Furthermore, we set the population of mean of class 1 to be a $p$-dimensional
vector of zeros. Next, we set class 2's population mean to a vector equaling 1/2
for the first 50 features and 0 after ($\mu_j = 1/2$ if $j \le 50$ and $\mu_j =
0$ if $j > 50$). Similarly, we set class 3's population mean to a vector equaling -1/2
for the first 50 features and 0 after ($\mu_j = -1/2$ if $j \le 50$ and $\mu_j =
0$ if $j > 50$). The three populations differ in their mean vectors in the first
50 features corresponding to the first two blocks. No difference in the means
occurs in the remaining blocks.

The block-diagonal simulation design considered is widely used within the
high-dimensional classification literature, including the studies by
\cite{Ramey:2013ji}, \cite{Witten:2011kc}, and \cite{Guo:2007te}. While the
design is indeed standard, the simulation configuration is based on an ideal
scenario where no artifacts, often observed in real data, are present. As a
result, we wished to investigate the effect of outliers on the high-dimensional
classifiers. To do this, we generalized the block-diagonal simulation
configuration by sampling from a $p$-dimensional multivariate
$\epsilon$-contaminated normal distribution, where $\epsilon in [0, 1]$. For
$\epsilon = 0$, we have the benchmark block-diagonal simulations considered in
the literature. As $\epsilon$ is increased, the number of outliers is increased
(on average). We consider the values of $\epsilon = 0, 0.05, 0.1, 0.15,
0.2$. For the contamination simulation, we define the probability density
function of the $k$th class to be

\begin{align*}
  p(\bm x | \omega_k) = (1 - \epsilon) N_p(\bm x | \bm \mu_k, \bm \Sigma) + \epsilon N_p(\bm x | \bm \mu_k, 100 * \bm \Sigma)
\end{align*}

From each of the $K=3$ populations, we sampled 25 training observations ($n_k =
25$ for all $k$) and 1000 test observations. After training each classifier on
the training data, we classified the test data sets and computed the proportion
of mislabeled test observations to estimate the classification error rate for
each classifier. Repeating this process 250 times, we computed the average of
the error-rate estimates for each classifier. We allowed the number of features
to vary from $p = 100$ to $p = 500$ in increments of 100 to examine the increase
in dimension while maintaining a small sample size.

>>>>

In discriminant analysis we wish to assign an unlabeled vector $\bm x \in
\mathbb{R}_{p \times 1}$ to one of $K$ unique, known classes by constructing a
classifier from $N$ training observations. Let $\bm x_i = (x_{i1}, \ldots,
x_{ip}) \in \mathbb{R}_{p \times 1}$ be the $i$th observation $(i = 1, \ldots,
N)$ with true, unique membership $y_i \in \{\omega_1, \ldots, \omega_K\}$. We
assume that $(\bm x_i, y_i)$ is a realization from a mixture distribution
$p(\bm x) = \sum_{k=1}^K p(\bm x | \omega_k) p(\omega_k)$, where $p(\bm x |
\omega_k)$ is the probability density function (PDF) of the $k$th class and
$p(\omega_k)$ is the prior probability of class membership of the $k$th
class. We further assume that each of $p(\omega_k)$ is equal, $k = 1, \ldots,
K$.

@@

In this section we compare the competing classifiers using the simulation
design from \cite{Guo:2007te}. This design is widely used within the
high-dimensional classification literature including the studies by
\cite{Ramey:2013ji} and \cite{Witten:2011kc}. The original design comprised two
$p$-dimensional multivariate normal populations with the common block-diagonal
covariance matrix
\begin{align}
  \bm\Sigma = \bm\Sigma^{(\rho)} \oplus \bm\Sigma^{(-\rho)} \oplus \ldots \oplus \bm\Sigma^{(\rho)},\label{eq:block-sigma}
\end{align}
where the $(i,j)$th entry of the block matrix $\bm\Sigma^{(\rho)} \in
\mathbb{R}_{25 \times 25}$ is
\begin{align*}
\bm\Sigma_{ij}^{(\rho)} = \{ \rho^{|i - j|} \}_{1 \le i,j \le 25}.
\end{align*}
The block-diagonal covariance structure in \eqref{eq:block-sigma} resembles
gene-expression data: within each block of pathways, genes are correlated, and
the correlation decays as a function of the distance between any two genes. The
autocorrelation $\rho$ alternates between a positive and a negative value from block
to block.

Although the design is indeed standard, the simulation configuration lacks
artifacts observed commonly in real data, such as skewness and extreme
outliers. As a result, we wished to investigate the effect of outliers on the
high-dimensional classifiers. To do this, we generalized the block-diagonal
simulation configuration by sampling from a $p$-dimensional multivariate
contaminated normal distribution.  Denoting the PDF of the $p$-dimensional
multivariate normal distribution by $N_p(\bm x | \bm \mu, \bm \Sigma)$, we
write the PDF of the $k$th class as
\begin{align}
  p(\bm x | \omega_k) = (1 - \epsilon) N_p(\bm x | \bm \mu_k, \bm \Sigma) + \epsilon N_p(\bm x | \bm \mu_k, \omega \bm \Sigma),\label{eq:contaminated-pdf}
\end{align}
where $\epsilon \in [0, 1]$ is the probability that an observation is contaminated (i.e.,
drawn from a distribution with larger variance) and $\omega > 1$ scales the
covariance matrix $\bm \Sigma$ to increase the extremity of outliers. For
$\epsilon = 0$, we have the benchmark block-diagonal simulation design from
\cite{Guo:2007te}. As $\epsilon$ is increased, the average number of outliers
is increased. In our simulation, we considered the values of $\epsilon = 0$,
$0.05$, $0.1$, $0.15$, and $0.2$. We also set $\omega = 100$.

We generated $K=3$ populations from \eqref{eq:contaminated-pdf} using the
covariance matrix $\bm \Sigma$ given in \eqref{eq:block-sigma}. We set the mean
vector of class 1 to $\bm \mu_1 = \bm 0_p$. Next, comparable to
\cite{Guo:2007te}, the first 50 features of $\bm \mu_2$ were 1/2, while the rest
were set to 0, i.e., $\bm \mu_2 = (\underbrace{1/2, \ldots, 1/2}_{50},
\underbrace{0, \ldots, 0}_{p - 50})$. For simplicity, we defined $\bm \mu_3 =
-\bm \mu_2$. The three populations differed in their mean vectors in the first
50 features corresponding to the first two blocks. No difference in the means
occurred in the remaining blocks.

From each of the $K=3$ populations, we sampled 25 training observations ($n_k =
25$ for all $k$) and 1000 test observations. After training each classifier on
the training data, we classified the test data sets and computed the proportion
of mislabeled test observations to estimate the classification error rate for
each classifier. Repeating this process 250 times, we computed the average of
the error-rate estimates for each classifier. We allowed the number of features
to vary from $p = 100$ to $p = 500$ in increments of 100 to examine the
classification accuracy as the feature dimension increased while maintaining a
small sample size. Furthermore, although \cite{Guo:2007te} consider $\rho =
0.9$, we considered $\rho = 0.1$, $0.5$, and $0.9$ to explore the effect of
autocorrelation on classification accuracy, especially given that the Tong,
Pang, and Guo classifiers assume independence among features.
